---
title: "CIPS"
author: "Oxana Rosca"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Do not use scientific notations
```{r}
options(scipen = 999)
```

```{r}
library(naniar)
library(readxl)
library(openxlsx)
library(dplyr)
library(ggplot2)
library(EnvStats)
```

# Check the "NEW CIPS for matching (N=2047)"
## Load "NEW CIPS for matching (N=2047)" clean dataset 
without children with incarcerated mothers
```{r}
cifs2047 <- as.data.frame(read_excel('CIFS_2047.xlsx'))
# # or
# cifs2047 <- as.data.frame(read_excel('NEW CIPS for matching (N=2047).xlsx'))

View(cifs2047)
```

## Remove Children with missing values on AgeM
```{r}
cifs2037 <- cifs2047[complete.cases(cifs2047$AgeM),]
```


## Missing Data Analysis for cifs2037
library(naniar)
```{r}
any_na(cifs2037)
```
How many? Percentage?
```{r}
n_miss(cifs2037)
prop_miss(cifs2037)
```
### 880 missing values on age variable. In the original dataset: 1102 
Get number of missings per variable (n and %)
```{r}
# Convert 'pillar_num' columns to numeric
cifs2037_missing <- miss_var_summary(cifs2037) %>%
  mutate(across(where(is.numeric), as.numeric))
# Display the data frame
cifs2037_missing
```
### Plot: Missingness on Variables 
```{r}
gg_miss_var(cifs2037)
```

## The dataset has 21 duplicate cases on the ID variable
```{r}
table(duplicated(cifs2037$idnumber),useNA = 'always')
```
## Decision: Reject the 'NEW CIPS for matching (N=2047).xlsx' and clean the original dataset.

# Clean and Organize the Original Dataset (N=2520).

## Load 2520 original dataset (all  children)
```{r}
cips_2520 <- as.data.frame(read_excel('CIPS_2520.xlsx'))
View(cips_2520)
```

## Create a variable of the row number in the original dataset
```{r}
cips_2520 <- cips_2520 %>%
  mutate(row.numberCIPS_2520 = row_number())
```

## Remove 1102 Children with missing values on ageM
```{r}
cifs_1418 <- cips_2520[complete.cases(cips_2520$AGEM),]
```

## The dataset has 163 duplicate cases on the ID variable
```{r}
table(duplicated(cifs_1418$idnumber),useNA = 'always')
```
### Missing Data Analysis for the original 1418 cases

```{r}
any_na(cifs_1418)
```
How many?
```{r}
n_miss(cifs_1418)
prop_miss(cifs_1418)
```
#### Get number of missings per variable (n and %)
```{r}
# Convert 'pillar_num' columns to numeric
cifs_1418_missing <- miss_var_summary(cifs_1418) %>%
  mutate(across(where(is.numeric), as.numeric))
# Write the modified data frame to an Excel file
write.xlsx(cifs_1418_missing, file = "cifs_1418_missing.xlsx")
# Display the data frame
cifs_1418_missing
```

#### Plot: Missingness on Variables 
```{r}
gg_miss_var(cifs_1418)
```

#### 2-12 missings per participant
```{r}
miss_case_summary(cifs_1418)
cifs_1418_missingPerCase<-miss_case_table(cifs_1418)
write.xlsx(cifs_1418_missingPerCase, file = "cifs_1418_missingPerCase.xlsx")
cifs_1418_missingPerCase
```

#### Missing Plot
Where are missings located?
```{r}
vis_miss(cifs_1418) + theme(axis.text.x = element_text(angle=80))
```
## Remove factor of incarcerated mother. Subset 1394 rows: 
139 with incarcerated fathers only and 1255 non-incarcerated parents
"{0, No Parent Incarcerated} {1, Father Incarcerated} 
{2, Mother Incarcerated} {3, Both}"
```{r}
table(cifs_1418$parincar, useNA = 'always')
```
Remove 24 rows:
```{r}
cifs_1394<-cifs_1418[cifs_1418$parincar<2,]
```

## Remove 157 Duplicated IDs

### 19 pairs of rows (38 rows) had identical IDs and the same year of program

```{r}
duplicates_program_ID <- cifs_1394 %>%
  group_by(idnumber, program) %>%
  filter(n() > 1)
write.xlsx(duplicates_program_ID, file = "cifs_1394_DuplicateRecords_SameProgram.xlsx")
print(duplicates_program_ID)
```

```{r}
table(duplicated(duplicates_program_ID$idnumber),useNA = 'always')
```
### 140 pairs of rows (380 rows) had identical IDs but different year of program 
There were one or two sets of three rows with identical id numbers.
```{r}
duplicates_program_ID2 <- cifs_1394 %>%
  group_by(idnumber) %>%
  filter(n() > 1) %>%
  ungroup()
write.xlsx(duplicates_program_ID2, file = "cifs_1394_DuplicateRecords_DifProgram.xlsx")
print(duplicates_program_ID2)
```

```{r}
table(duplicated(duplicates_program_ID2$idnumber),useNA = 'always')
```

## Remove 157 Duplicated IDs
See the reasoning for removal in "cips_2520_data cleaning.xlsx".
"DuplicateRecords_SameProgram" and "DuplicateRecords_DifProgram": 
Retain 157 green rows and remove 157 orange rows in the tab "Items to remove"
```{r}
row.numberCIPS_2520_to_remove <- c(34, 55, 58, 72, 74, 76, 78, 80, 82, 84, 87, 
    90, 93, 96, 98, 100, 104, 106, 108, 132, 134, 136, 138, 141, 143, 145, 147, 
    149, 151, 154, 156, 158, 161, 163, 165, 167, 169, 171, 173, 175, 193, 216, 
    225, 242, 252, 259, 262, 273, 324, 335, 337, 355, 359, 384, 437, 444, 446, 
    452, 453, 455, 458, 461, 463, 465, 467, 470, 471, 473, 475, 479, 482, 484, 
    487, 496, 507, 510, 516, 540, 543, 547, 550, 552, 554, 556, 565, 580, 589, 
    597, 599, 602, 607, 616, 637, 651, 654, 657, 664, 674, 676, 678, 680, 683, 
    685, 687, 689, 691, 694, 696, 699, 701, 704, 708, 711, 717, 719, 721, 723, 
    725, 728, 730, 733, 736, 820, 822, 845, 848, 850, 852, 854, 856, 858, 874, 
    879, 901, 917, 920, 967, 969, 972, 975, 977, 979, 981, 986, 988, 998, 1002, 
    1016, 1018, 1020, 1025, 1032, 1054, 1087, 1097, 1130, 1755)
cifs_1237 <- cifs_1394 %>%
  filter(!(row.numberCIPS_2520 %in% row.numberCIPS_2520_to_remove))
```

## Adjust the missing and wrong values
See the reasoning in "cips_2520_results" in the tabs "DuplicateRecords_DifProgram" 
and "DuplicateRecords_SameProgram"

### Change 'program' value from 2 to 1 for the row.numberCIPS_2520=272 (idnumber = 259)
```{r}
cifs_1237[cifs_1237$row.numberCIPS_2520 == 272, "program"] <- 1
```

### Change 'program' value from 1 to 2 for 8 specified row.numberCIPS_2520
id numbers are 811	854	987	1012	1018	1183	1186	1209
```{r}
row.numberCIPS_2520_to_changeProgram <- 
                        c(684,737,851,875,880,999,1003,1026)
cifs_1237 <- cifs_1237 %>% 
  mutate(program = ifelse(row.numberCIPS_2520 %in% row.numberCIPS_2520_to_changeProgram, 2, program))
```

### Change "I-to-N ratio" variable's name to avoid the space
```{r}
names(cifs_1237)[names(cifs_1237) == "I-to-N ratio"] <- "I_to_N_ratio"
```

### Incert missing values=0 into "income" and	"I_to_N_ratio" for the specified row.numberCIPS_2520
id numbers are 56, 76, 138, 525, 1279.  
```{r}
row.numberCIPS_2520_to_changeIncome <- c(56, 75, 164, 445, 1088)
cifs_1237 <- cifs_1237 %>% 
  mutate(income = ifelse(row.numberCIPS_2520 %in% row.numberCIPS_2520_to_changeIncome, 0, income))
cifs_1237 <- cifs_1237 %>% 
  mutate(I_to_N_ratio = ifelse(row.numberCIPS_2520 %in% row.numberCIPS_2520_to_changeIncome, 0, I_to_N_ratio))
```

## The dataset has no duplicate cases on the ID variable
```{r}
table(duplicated(cifs_1237$idnumber),useNA = 'always')
```

## Missing Data Analysis for 1237 cases (clean dataset)

```{r}
any_na(cifs_1237)
```
How many?
```{r}
n_miss(cifs_1237)
prop_miss(cifs_1237)
```
### 880 missing values on age variable. In the original dataset: 1102 
Get number of missings per variable (n and %)
```{r}
# Convert 'pillar_num' columns to numeric
cifs_1237_missing <- miss_var_summary(cifs_1237) %>%
  mutate(across(where(is.numeric), as.numeric))
# Display the data frame
cifs_1237_missing
```
### Plot: Missingness on Variables 
```{r}
gg_miss_var(cifs_1237)
```

#### 0-2 missings per participant
```{r}
miss_case_summary(cifs_1237)
cifs_1237_missingPerCase<-miss_case_table(cifs_1418)
write.xlsx(cifs_1237_missingPerCase, file = "cifs_1418_missingPerCase.xlsx")
cifs_1237_missingPerCase
```

### Missing Plot
Where are missings located?
```{r}
vis_miss(cifs_1237) + theme(axis.text.x = element_text(angle=80))
```

# Getting the Analytical Sample (N=1181)

## Remove 4 cases with missing values on family size and 52 cases with missing values on income 
```{r}
cifs_1233 <- cifs_1237[complete.cases(cifs_1237$`Family size`),]
AnSamp_cifs_1181 <- cifs_1233[complete.cases(cifs_1233$income),]
```

#### Analytic Sample Missing Plot
Where are missings located?
```{r}
vis_miss(AnSamp_cifs_1181) + theme(axis.text.x = element_text(angle=80))
```

# Check the summary of the dataset and save it as an Excel file
```{r}
AnSamp_cifs_1181_summary<-summaryFull(AnSamp_cifs_1181)
AnSamp_cifs_1181_summary
write.xlsx(AnSamp_cifs_1181_summary, file = "Analytic Sample cifs_1181 summary.xlsx", rowNames = TRUE)
write.xlsx(AnSamp_cifs_1181, file = "Analytic Sample cifs_1181 Data.xlsx", rowNames = F)
```

# There are 121 children with incarcerated fathers and 1060 children without incarcerated parents.
```{r}
table(AnSamp_cifs_1181$parincar, useNA = 'always')
```
